A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
re-wrote intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@484 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Color is an integral and omnipresent element of many statistical graphics and data
visualizations. Therefore, colors should be carefully chosen to support the information
displayed [@color:Tufte:1990, @color:Brewer:1999, @color:Ware:2004, @color:Wilkinson:2005, @color:Wilke:2019].
However, until relatively recently many software packages have been using color
palettes derived from simple RGB (red-green-blue)
color combinations such as the RGB "rainbow" (or "jet") color palette
with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.

To address these problems, many improved color palettes with better perceptual
properties have been suggested in the literature
[@color:Harrower+Brewer:2003, @color:Zeileis+Hornik+Murrell:2009, @color:Smith+VanDerWalt:2015, @color:Crameri:2018]
and also made available in standard software packages including various
packages for \proglang{R} [@color:R] such as 
\pkg{RColorBrewer} [@color:RColorBrewer],
\pkg{rcartocolor} [@color:rcartocolor],
\pkg{viridis} [@color:viridis],
\pkg{wesanderson} [@color:wesanderson],
or \pkg{scico} [@color:scico] among many others.
In packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
many of these palettes are collected and provided with a unified interface.
Most of these palettes, however, are pre-existing palettes, stored as a
limited set of colors and interpolated as necessary. And even if specific
algorithms have been used in the first construction of the palettes, these
are often not reflected in the software implementations.

In contrast, the \pkg{colorspace} package [@color:colorspace] adopts a perceptually-based
approach -- the so-called HCL (hue--chroma--luminance) space -- and
establishes a broad range of color palettes through general and adaptable
trajectories in HCL space. In addition to constructing the palettes the HCL
space is also employed to visualize and assess the properties of a given
palette (beyond simple color swatches). The reason for choosing the HCL space
for all of these tools is that its three axes match those of the human visual system very well:
Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.

@@

The \pkg{colorspace} package provides three types of palettes based on the HCL model:

@@

To aid choice and application of these palettes there are: scales for use
with \pkg{ggplot2} [@color:ggplot2]; \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration;
visualizations of palette properties; accompanying manipulation utilities
(like desaturation, lighten/darken, and emulation of color vision deficiencies).

FIXME: More comparison with/discussion of other palette packages.
the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
and can be approximated reasonably well in \pkg{colorspace}
(see Section\ \ref{sec:approximations}).
A broad range of packages that provide color palettes that are "in harmony" as
they are extracted and clustered from movies, paintings, etc. However, many of these
are not so suitable for scientific visualization as they are not clearly 
qualitative/sequential/diverging. An influential package following such an
approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
[@color:paletteer] package aggregates these and provides them in a unified
user interface.
The \pkg{pals} package also provides a large number of palettes,
but some are just pre-existing palettes, and where the palettes are new
there is no description of a specific algorithm that was used to generate
them.  
Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
coordinates, along with visualization and assessment of palettes (beyond
color swatches).

This paper is organized as follows...

The \pkg{colorspace} package provides a broad toolbox for selecting individual
colors or color palettes, manipulating these colors, and employing
them in various kinds of visualizations.

At the core of the package there are various utilities for computing with
color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
representations of color to each other. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in \proglang{R}). For completeness further standard color models are included as well in the package:
`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
`HSV()`.

@@

At the core of the \pkg{colorspace} package are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other [@color:Ihaka:2003]. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes
[@color:Wiki+Webcolors], e.g., in HTML but also in \proglang{R}.
For completeness further standard color models are included as well in the package.
The connections are illustrated in Figure\ \ref{fig:color-spaces}.
Color models that are (or try to be) perceptually-based are displayed with circles
and models that are not are displayed with rectangles.

@@

Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for $H = 245$. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

@@

* **Palette constructor:** `choose_palette()` or `hclwizard()` or `hcl_wizard()`.
* **Color picker:** `choose_color()` or equivalently `hcl_color_picker()`.
* **Color vision deficiency emulator:** `cvd_emulator()`.

@@

Optionally, the active palette can be illustrated by using a `specplot()` (see Section\ \ref{sec:specplot}),
`hclplot()` (see Section\ \ref{sec:hclplot}), or `demoplot()`
(see Section\ \ref{sec:demoplot}), and assessed using emulation
of color vision deficiencies (see Section\ \ref{sec:color_vision_deficiency}).
To facilitate generation of palettes for black/dark backgrounds, a "dark mode"
of the GUIs is also available:

@@

In principle, a similar approach for lightening and darkening colors can be employed as for desaturation
above. The colors can simply be transformed to HCL space and then the luminance can either be decreased
(turning the color darker) or increased (turning it lighter) while preserving the hue and chroma coordinates.
This strategy typically works well for lightening colors, although in some situations the result can be
rather colorful. Conversely, when darkening rather light colors with little chroma, this can result
in rather gray colors.

@@

Regardless of the chosen color space, the adjustment of the `L` component can occur by two methods, relative
(the default) and absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
`L * (1 - amount)` (relative). See `?lighten` and `?darken` for more details.

>>>>

Color is an integral and omnipresent element of many statistical graphics and data
visualizations. Therefore, colors should be carefully chosen to support all viewers in
accessing the information displayed [@color:Tufte:1990; @color:Brewer:1999; @color:Ware:2004; @color:Wilkinson:2005; @color:Wilke:2019].
However, until relatively recently many software packages have been using color
palettes derived from simple RGB (red-green-blue)
color combinations such as the RGB "rainbow" (or "jet") color palette
with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.

To address these problems, many improved color palettes with better perceptual
properties have been suggested in the literature over the last one or two decades
[@color:Harrower+Brewer:2003; @color:Zeileis+Hornik+Murrell:2009; @color:Smith+VanDerWalt:2015; @color:CARTO; @color:Crameri:2018].
Many systems for systems for statistical and scientific computing provide
infrastructure for such color palettes, e.g., for \proglang{R} [@color:R] the
list of useful packages encompasses:
\pkg{RColorBrewer} [@color:RColorBrewer],
\pkg{viridis} [@color:viridis],
\pkg{rcartocolor} [@color:rcartocolor],
\pkg{wesanderson} [@color:wesanderson],
or \pkg{scico} [@color:scico] among many others.
Furthermore, packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
collect many of the proposed palettes in combination with a unified interface.
Most of these palettes, however, are pre-existing palettes, stored as a
limited set of colors and interpolated as necessary. And even if specific
algorithms have been used in the initial construction of the palettes, these
are often not reflected in the software implementations.

@@

The \pkg{colorspace} package [@color:colorspace] adopts a somewhat different
approach, utiliting the perceptually-based HCL (hue--chroma--luminance) color space
[@color:Wiki+HCL]. The three axes of the HCL space match those of the human visual
system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.
Thus, based on this color model the \pkg{colorspace} package can derive general and adaptable
strategies for color palettes; manipulate individual colors and color palettes;
and assess and visualize the properties of color palettes (beyond simple color swatches).
Specifically, \pkg{colorspace} provides three types of palettes based on the HCL model:

@@

A broad collection of prespecified palettes are shipped in the package, existing
palettes can be easily tweaked, and new or adapted palettes registered. The prespecified
palettes include suitable HCL color choices that closely approximate of most palettes
from packages \pkg{RColorBrewer}, \pkg{rcartocolor}, and \pkg{viridis} by using only a
small set of hue, chroma, and luminance parameters.

To aid choice and application of these palettes there are
(a) scales for use with \pkg{ggplot2} [@color:ggplot2],
(b) \pkg{shiny} [@color:shiny] [and \pkg{tcltk}, @color:R] apps for interactive exploration,
(c) visualizations of palette properties, and
(d) accompanying manipulation utilities (like desaturation, lighten/darken, and emulation of color vision deficiencies).

The remainder of the manuscript is organized as follows:
Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
that help assess the suitability of colors for colorblind viewers.
Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
before Section\ \ref{sec:summary} concludes the manuscript.

@@

At the core of the \pkg{colorspace} package are various utilities for computing with
color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
representations of color to each other [@color:Ihaka:2003]. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes
[@color:Wiki+Webcolors], e.g., in HTML but also in \proglang{R}.
For completeness further standard color models are included as well in the package.
The connections are illustrated in Figure\ \ref{fig:color-spaces}.
Color models that are (or try to be) perceptually-based are displayed with circles
and models that are not are displayed with rectangles.

@@

Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for $H = 245$ because of
the distorted shape of the HCL space. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

@@

* _Palette constructor:_ `choose_palette()` or `hclwizard()` or `hcl_wizard()`.
* _Color picker:_ `choose_color()` or equivalently `hcl_color_picker()`.
* _Color vision deficiency emulator:_ `cvd_emulator()`.

@@

Optionally, the active palette can be illustrated by using a `specplot()` (see Section\ \ref{sec:specplot}),
`hclplot()` (see Section\ \ref{sec:hclplot}), or `demoplot()`
(see Section\ \ref{sec:demoplot}), and assessed using emulation
of color vision deficiencies (see Section\ \ref{sec:color_vision_deficiency}).
To facilitate generation of palettes for black/dark backgrounds, a "dark mode"
of the GUIs is also available.

@@

In principle, a similar approach for lightening and darkening colors can be employed as for desaturation
above. The colors can simply be transformed to HCL space and then the luminance can either be decreased
(turning the color darker) or increased (turning it lighter) while preserving the hue and chroma coordinates.
This strategy typically works well for lightening colors, although in some situations the result can be
somewhat too colorful. Conversely, when darkening rather light colors with little chroma, this can result
in rather gray colors.

@@

Regardless of the chosen color space, the adjustment of the `L` component by a certain `amount` can occur by two methods, relative
(the default) or absolute. For example for darkening these either use `L - 100 * amount` (absolute) or
`L * (1 - amount)` (relative). See `?lighten` and `?darken` for more details.
